---
title: "Database Connections"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Connecting to Databases from R

> Patterned after http://db.rstudio.com/

This walkthrough is for MacOS X users. Windows users may be able to achieve 
similar results through the use of Chocolatey, though research into the 
specific packages under Windows that match what homebrew installs is 
required.

These connectivity examples use DBI-compatible interfaces. DBI is a database 
interface specification that allows any compliant tool to communicate with 
databases without knowledge of the underlying database technology. Using 
DBI tooling allows both direct SQL queries through raw connections as well as 
tight integration with the tidyverse through the spiffy dbplyr package.

Database drivers must be written with DBI interfaces. Both native drivers and the 
generic (DBI-compliant) odbc connections are demonstrated. For Oracle and MS 
SQL Server, odbc is the only (free) connectivity option.

If you are only connecting to a single database type, the driver you choose is 
largely a matter of personal preference. If connecting to multiple different 
database types, using odbc will provide a more consistent interface. 
Connections made using odbc will appear in the connections tab of the RStudio 
IDE.

## Install ODBC and Database Drivers

```{shell odbc_install, eval=FALSE}
# install base odbc framework
brew install unixodbc

# Microsoft SQL Server, mkaing it available for odbc and with Kerberos support
brew install freetds --with-unixodbc --with-krb5
  
# PostgreSQL ODBC Drivers
brew install psqlodbc

# Maria/MySQL ODBC drivers
brew install mariadb-connector-c
brew install mariadb-connector-odbc

# SQLite ODBC Drivers
brew install sqliteodbc
```

### Oracle Client Install

The Oracle client install is complicated. `r emo::ji("frowning")`

Download the Basic client and the ODBC drivers via  
http://www.oracle.com/technetwork/topics/intel-macsoft-096467.html. A (free) 
Oracle account is required. Place both ZIP files in your ~/Downloads directory.

```{bash, eval=FALSE}
sudo unzip -o ~/Downloads/instantclient-basic-macos.x64-12.2.0.1.0-2.zip -d /opt/ora12
sudo unzip -o ~/Downloads/instantclient-odbc-macos.x64-12.2.0.1.0-2.zip -d /opt/ora12

# patch the Oracle ODBC install script to fix an incorrectly specified library
# extension (.so -> .dylib)
sudo patch ./oracle_odbc_update.patch /opt/ora12/instantclient_12_2/odbc_update_ini.sh > patch.diff
sudo /opt/ora12/instantclient_12_2/odbc_update_ini.sh /usr/local

ln -s `pwd`/libclntsh.dylib.12.1 /usr/local/lib
ln -s `pwd`/libclntshcore.dylib.12.1 /usr/local/lib
```


### Configure ODBC Drivers

Set `/usr/local/etc/odbcinst.ini` to the following to allow easy refernce 
to the database drivers by name rather than by full file path:

```
[SQLite]
Driver = /usr/local/lib/libsqlite3odbc.dylib

[PostgreSQL]
Driver = /usr/local/lib/psqlodbcw.so

[MySQL]
Driver = /usr/local/lib/libmysqlclient.dylib

[MariaSQL]
Driver = /usr/local/lib/libmaodbc.dylib

[SQL Server]
Driver = /usr/local/lib/libtdsodbc.so

[Oracle 12c]
Description     = Oracle ODBC driver for Oracle 12c
Driver          = /usr/local/lib/libsqora.dylib.12.2
Setup           =
FileUsage       =
CPTimeout       =
CPReuse         =
```

## Configure R packages

Install all the relevant R packages. We use DBI as the primary database 
interface, with odbc for ODBC connectivity. Native drivers for PostgreSQL,
MariaDB/MySQL, and SQLite are also installed. In production, you can 
choose either native or ODBC drivers as you see fit.

```{r, eval=FALSE}
install.packages("DBI")         # DBI framework
install.packages("odbc")        # odbc framework

# Native drivers
install.packages("RPostgres")
install.packages("RMariaDB")    # covers MariaDB and MySQL
install.packages("RSQLite")
```
