---
title: "Database Connections"
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    toc: yes
---

```{r setup}
library(DBI)
library(tidyverse)
library(dbplyr)
```

# Establish a Connection

The following sections show how to create connection to various flavors of 
databases.

## SQLite
```{r sqlite_connect}
con <- dbConnect(RSQLite::SQLite(),  ":memory:")
```

## Postgres
```{bash launch_postgres, eval=FALSE}
docker run --rm --name rstats-postgres -e POSTGRES_PASSWORD=mysecretpassword -d -p 5432:5432 postgres 
```

Connect to PostgreSQL via DBI or ODBC.

```{r connect_postgres}
con <- dbConnect(RPostgres::Postgres(),
        host      = "127.0.0.1",
        password  = 'mysecretpassword', 
        user      = 'postgres') 

con <- dbConnect(odbc::odbc(), 
                 Driver   = "PostgreSQL",
                 Database = "postgres",
                 PWD      = 'mysecretpassword', 
                 UID      = 'postgres', 
                 Server   = "127.0.0.1", 
                 port     = 5432)
```


## MariaDB/MySQL

Launch either a MariaDB or MySQL docker container.

```{bash, eval=FALSE}
docker run --rm --name rstats-mariadb -e MYSQL_ROOT_PASSWORD=mysecretpassword -d -p 3306:3306 mariadb
docker run --rm --name rstats-mysqldb -e MYSQL_ROOT_PASSWORD=mysecretpassword -d -p 3307:3306 mysql
```

Connect to MySQL

```{r mysql}
con <- dbConnect(RMariaDB::MariaDB(),
          dbname    = "mysql",
          host      = "127.0.0.1", 
          port      = 3307,
          user      = 'root', 
          password  = 'mysecretpassword') 

con <- dbConnect(odbc::odbc(), 
          Driver   = 'MariaSQL',
          Server   = "127.0.0.1", 
          Port     = 3307,
          Database = 'mysql',
          Uid      = 'root', 
          Pwd      = 'mysecretpassword')
```

Connect to MariaDB

```{r maria}
con <- dbConnect(RMariaDB::MariaDB(),
          dbname    = "mysql",
          host      = "127.0.0.1", 
          port      = 3306,
          user      = 'root', 
          password  = 'mysecretpassword') 

con <- dbConnect(odbc::odbc(), 
          Driver   = 'MariaSQL',
          Server   = "127.0.0.1", 
          Port     = 3306,
          Database = 'mysql',
          Uid      = 'root', 
          Pwd      = 'mysecretpassword')
```

## MS SQL Server

```{bash launch_mssql, eval=FALSE}
docker run --rm --name rstats-mssql -e 'ACCEPT_EULA=Y' -e 'SA_PASSWORD=myS3cretPassword' -p 1433:1433 -d microsoft/mssql-server-linux
```

```{r connect_mssql}
con <- dbConnect(odbc::odbc(), 
          Driver   = 'SQL Server',
          Server   = "127.0.0.1", 
          Port     = 1433,
          Database = 'master',
          Uid      = 'sa', 
          Pwd      = 'myS3cretPassword')
```

## Oracle

Per https://serverfault.com/a/51304

* SID: identifies the database instance (database name + instance number). So 
if your database name is somedb and your instance number is 3, then your SID 
is somedb3.

* Instance name: same as SID

* DB Name: Name of the database (database can be shared b/t multiple instances)

* DB Domain:  Usually the same as your company domain (somecompany.com)

* Global (Database) Name: Database name + database domain (somedb.somecompany.com)

* Service Name: A "connector" to one or more instances. It is often useful to 
create additional service names in a RAC environment since the service can be 
modified to use particular SIDs as primary or secondary connections, or to not 
use certain SIDs at all.

* Service Alias: An alias to the service name (just like a CNAME, etc). Say 
you make your service name something meaningful to the dba, but perhaps it's a 
bit esoteric. Create a service alias and name it something that will be 
meaningful to the user.

Set tnsnames.ora to the following:
```
DB1 =
 ( DESCRIPTION =
   (ADDRESS_LIST =
     (ADDRESS =
       (PROTOCOL = TCP)
       (Host = myoracledb.myorg.tld)
       (Port = 15033)
     )
   )
   (CONNECT_DATA =
    (SID = Special)
   )
 )
```
 
Ensure that TNS_ADMIN is pointed to the directory where you have tnsnames.ora 
located. For me, that is `/Users/davidski`. This can be set in the .Renviron in 
the project directory.

Set odbc.ini to

```
[db1]
Description = DB1 Oracle Database
Driver = Oracle 12c
Servername = DB1
```

```{r oracle}
#con <- dbConnect(odbc::odbc(), dsn = "db1")
con <- dbConnect(odbc::odbc(), dsn = "db1", 
                 Uid = rstudioapi::askForPassword("DB1 Username"),
                 Pwd = rstudioapi::askForPassword("DB1 password"))
```

# Using Database Connections

With a valid connection to a database (`con` variable in this example), we 
can now work with our database in a variety of SQL and R-friendly ways.

## Testing DBI interface


```{r}
dbListTables(con)
dbWriteTable(con, "mtcars", mtcars)
dbListTables(con)

dbListFields(con, "mtcars")
dbReadTable(con, "mtcars")

# You can fetch all results:
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
dbFetch(res)
dbClearResult(res)

# Or a chunk at a time
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
while(!dbHasCompleted(res)){
  chunk <- dbFetch(res, n = 5)
  print(nrow(chunk))
  print(chunk)
}
# Clear the result
dbClearResult(res)

# Disconnect from the database
dbDisconnect(con)
```

## Testing dbplyr interface

Load data into the DB
```{r create_flights}
copy_to(con, nycflights13::flights, "flights", 
        overwrite = TRUE,
        temporary = FALSE,
        indexes = list(
          c("year", "month", "day"),
          "carrier",
          "tailnum",
          "dest")
)

flights_db <- tbl(con, "flights")
flights_db
```

Perform a simple count

```{r}
flights_db %>% tally()
```

Perform a more complex dplyr pipe

```{r}
tailnum_delay_db <- flights_db %>% 
  group_by(tailnum) %>%
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  ) %>% 
  arrange(desc(delay)) %>%
  filter(n > 100)

tailnum_delay_db %>% collect()
```

```{sql raw_sql, connection=con, eval = FALSE}
SELECT * from flights WHERE MONTH = 1
```
